PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-12 IEEEtran/annex.tex
Add annex

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,citecolor=blue!75!black,filecolor=black,urlcolor=blue!75!black}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage[frozencache=true,cachedir=minted-cache]{minted} 
\usepackage[htt]{hyphenat}      
\usepackage{subcaption}
\usepackage{graphicx,import}
\usepackage{booktabs}

% \usepackage{subfigure}

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1} #2\color{black}\xspace}
\newcommand{\gkmod}[2]{\color{purple}\sout{#1} #2\color{black}\xspace}

\newcommand{\Yohan}[1]{\color{green!75!black}\textbf{Yohan:} #1\color{black}\xspace}
\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\na}[0]{\cellcolor{gray!25}}
\newcommand{\s}[1]{\cellcolor{cyan!25}#1}
\newcommand{\scross}[0]{\ding{53}~}
\newcommand{\svalid}[0]{\ding{51}~}
\newcommand{\swarn}[0]{?~}
\newcommand{\pytracer}[0]{PyTracer\xspace}

\setlength{\arrayrulewidth}{0.1pt}

\include{href}

\usepackage{tabularx} % provides tabularx to adjust table column sizes to page size
% define thickline for use in tables
\makeatletter
\newcommand{\thickhline}{%
    \noalign {\ifnum 0=`}\fi \hrule height 1pt
    \futurelet \reserved@a \@xhline
}
\newcolumntype{"}{@{\hskip\tabcolsep\vrule width 1pt\hskip\tabcolsep}}
\makeatother

\lstdefinestyle{customPython}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  xleftmargin=\parindent,
  language=Python,
  showstringspaces=false,
  basicstyle=\scriptsize\ttfamily,
  keywordstyle=\bfseries\color[rgb]{0.580, 0.000, 0.827},
  %{purple!40!lightgray},
  commentstyle=\itshape\color{green!40!black},
  identifierstyle=\bfseries\color{cyan!75!black},
  stringstyle=\color{orange},
  deletekeywords={double,float},
  classoffset=1, % starting new class
  otherkeywords={double,float},
  morekeywords={double,float},
  keywordstyle=\bfseries\color{green!55!black},
  classoffset=0
}

\lstdefinestyle{customC}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  xleftmargin=\parindent,
  language=C,
  showstringspaces=false,
  basicstyle=\scriptsize\ttfamily,
  keywordstyle=\bfseries\color[rgb]{0.580, 0.000, 0.827},
  %{purple!40!lightgray},
  commentstyle=\itshape\color{green!40!black},
  identifierstyle=\bfseries\color{cyan!75!black},
  stringstyle=\color{orange},
  deletekeywords={double,float},
  classoffset=1, % starting new class
  otherkeywords={double,float},
  morekeywords={double,float},
  keywordstyle=\bfseries\color{green!55!black},
  classoffset=0
}

\begin{document}

\makeatletter
\let\orig@lstnumber=\thelstnumber
\newcommand\lstsetnumber[1]{\gdef\thelstnumber{#1}}
\newcommand\lstresetnumber{\global\let\thelstnumber=\orig@lstnumber}
\makeatother

\title{PyTracer: Automatically profiling numerical instabilities in Python}
\author{Yohan Chatelain$^1$, Nigel Yong$^1$,  Gregory Kiar$^2$,  Tristan Glatard$^1$\\
$^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
$^2$Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
\date{\textit{This work has been submitted to the IEEE for possible publication. 
Copyright may be transferred without notice, after which this version may no 
longer be accessible.}}
\maketitle

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To fill this gap, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer transparently instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard. We designed \pytracer to be agnostic to numerical noise model , allowing for tool evaluation through Monte-Carlo Arithmetic, random rounding, random data perturbation, or structured noise for a particular application. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in both SciPy and Scikit-learn, two dominant Python libraries for mathematical modeling. Through these evaluations, we demonstrate \pytracer as a scalable, automatic, and generic framework for numerical profiling in Python.
% Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To address this issue, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard.
% \pytracer is designed to work with any type of numerical noise model, including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. Finally, \pytracer slows down executions by only $3.5 \times$ on average.
% \tristan{Add one sentence to summarize the findings} 
\end{abstract}

\section{Introduction}

The scientific Python ecosystem has become a central component of data
analyses in recent years, owing to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications each year, and serve as a reference set of high-quality open-source core scientific libraries. Researchers have built numerous domain-specific software tools from this ecosystem, leveraging Python's simplicity and flexibility.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by data noise, software and hardware updates, or parallelization have potential to introduce substantial deviations in final results, leading to potentially different scientific conclusions, ultimately threatening the reliability of computational analyses. To prevent this, numerical stability analyses need to be conducted systematically, however, no practical tool currently exists to conduct such analyses for Python programs.

We present PyTracer, a numerical stability profiling and visualization tool for Python programs.
\pytracer adopts a dynamic approach that evaluates numerical stability using program execution traces, and is applicable to large programs without manual intervention. The transparent nature of \pytracer is particularly valuable in contrast to theoretical approaches based on condition numbers or backward error analysis, which require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques hardly scale to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) generating numerically-perturbed executions, (2) tracing floating-point computations along program executions, and (3) visualizing stability evaluations. \pytracer addresses these requirements with (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

This manuscript presents the design of \pytracer and demonstrates its potential in various use cases. 
Through stability evaluations of the
SciPy and scikit-learn libraries, we demonstrate \pytracer as a practical solution to review the numerical stability of extensive code bases. 
%All information was obtained through \pytracer, visualized separately for improved customizability.
%\tristan{this is a weird place to mention this ^^}

\section{\pytracer design}

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}. First, \pytracer uses \textit{monkey patching} to replace the functions called by the application with a wrapper that saves their arguments and returned values in a trace file.
% \gkmod{}{Q: This is conceptually similar to a decorator, no? If so, perhaps it is worth mentioning} \Yohan{Conceptually, monkey patching is the ability to dynamically change a class or a module. So it's more powerful than the decorators. But yes we use monkey-patching to automatically decorate the functions.}
Second, it executes the application multiple times with noise injection, and aggregates the resulting trace files to compute summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

% \tristan{mention monkey patching}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow
    % \gkmod{}{Note: replace "Execution" with "Multiple Independent Executions", and maybe move it to be between the gears and the pkl files.}
    }
    \label{fig:workflow}
\end{figure}

\subsection{Python code instrumentation}

%\tristan{``Instrumentation" should appear in Figure 1}

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved Python methods.
However, some attributes are not writable. In this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches. In addition, this technique is scalable and applicable in read-only environments such as Singularity containers or HPC clusters.

\subsubsection{Intercepting module import}

Python loads modules through an import mechanism triggered by the \texttt{import} keyword that involves two objects: \texttt{finders} and \texttt{loaders}. \texttt{Finders} search for packages (or 
namespaces\footnote{Python distinguishes between \texttt{regular} and \texttt{namespace} packages.
A regular package is a directory that contains a \_\_init\_\_.py file and potentially subdirectories (sub-packages) 
that contain themselves a \texttt{\_\_init\_\_.py} file and so on recursively. 
The package hierarchy follows those of the directory. 
Namespace packages introduced in Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not contain an
\_\_init\_\_.py file and allow for flexible directory structures. Hence, parts of the package can be located in zip files, on the network, or in separated directories. There are no functional differences between both types of packages.}) on the local storage --- such as standard packages installed through the pip package manager --- or on a distant server.
\texttt{Finders} do not load modules. Instead, they return a specification (\texttt{spec} object) that includes 
information on where to find and how to load modules.
\texttt{Loaders} create module objects, initialize them, and fill import-related attributes 
such as \texttt{\_\_spec\_\_}. 
\texttt{Loaders} then execute modules and populate their namespace. Finally, modules are bound to their import name in the \texttt{sys.modules} dictionary.

Python supports custom \texttt{finder} classes registered in the \texttt{sys.meta\_path} list.
When Python encounters an \texttt{import} statement, it first looks for an existing binding in \texttt{sys.modules} and then iterates over the \texttt{finder} classes in the \texttt{sys.meta\_path} list until it finds the module. \pytracer adds a custom \texttt{finder} class at the head of \texttt{sys.meta\_path} that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module, then copies it as a new instance of the \texttt{ModuleType} class. It then calls the appropriate instrumentation function for each attribute depending on its type (function, class, or instances) and replaces the original module in the \texttt{sys.modules} map with the instrumented version. Finally, it updates all existing references to the original module contained in the global symbol table.
As a result, the application will transparently call the instrumented module.

% \tristan{is the non-instrumented version ever added to sys.modules? If so, could this lead to race conditions? If not, replace with ``adds the instrumented module to the sys.modules map"}. 
% \Yohan{Yes, it replaces the original module. You can't have a race-condition since sys.modules is a dict with only one value for a given key. The trick is that the wrapped module keeps a reference to the original module, it's just that this reference is invisible for the others object since it's not in sys.modules and globals. By the way, Pytracer needs to update all possible references that original objects may hold, like functions in \_\_dict\_\_ attributes or locals references in \_\_globals\_\_ for example.}

% \pytracer intercepts the modules loaded with the \texttt{Finder} and \texttt{Loader} mechanisms \tristan{is it restrictive? Are there other mechanisms?}.
% The Finder searches the loader of a module that is being imported while the 
% Loader loads and initializes the module. \pytracer adds \tristan{overrides?} a new Finder and Loader on top of the default ones
% to intercept imports and to instrument modules selected by the user. This mechanism allows returning 
% the original module in the instrumentation code and returning the instrumented module in the targeted application
% without modifying its actual code.

\subsubsection{Instrumenting module functions}

Listings~\ref{fig:wrapper_creation} and~\ref{fig:generic_wrapper} show \pytracer's function instrumentation mechanism.
For each function to instrument, \pytracer's \texttt{loader} class creates a wrapper function, dynamically compiles it with the \texttt{compile} builtin function, and substitutes it for the original module function.  Although simple, this technique does not apply to callable class instances, i.e., class instances that have a \texttt{\_\_call\_\_} attribute and are not of type \texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify the type of the class instance to \texttt{FunctionType}, which could cause syntax and semantic errors, and 2) mask class attributes, leading to \texttt{AttributeError} exceptions.
\pytracer overrides the \texttt{\_\_call\_\_} attribute with the wrapper function when possible to overcome this issue.
When the \texttt{\_\_call\_\_} attribute is not writable,  Pytracer does not instrument the class and returns a warning.

% \tristan{add docstrings to all Python code examples and clarify captions accordingly: you don't need to explain the meaning of each attribute in the caption if it's in the docstring. The caption can focus on the main functionality, reusing the text currently in 3.1.5.}
% \Yohan{TODO}

% Listing~\ref{fig:wrapper_creation} shows how functions are instrumented.
% This function takes as inputs the information about the function to instrument (module name and full qualified name),
% the function itself, and the actual name of the function. It then creates a wrapper around the actual wrapper
% (\texttt{generic\_wrapper}) that will save the values in trace files. Instead of passing the function, we pass the function's identifier, which is a unique 64-bits integer that references each Python object. This identifier is available
% with the \texttt{id} builtin function. it allows to get rid of aliases and to avoid instrumenting functions twice.  \tristan{Summarize this paragraph in the figure caption}

% Listing~\ref{fig:generic_wrapper} shows what the instrumented code does. 
% It first unpacks the function's information to get the function's identifier to retrieve the original function.
% Then arguments are bound by creating a mapping from positional and keyword arguments to parameters, avoiding conflicts and mispositioning. It also allows getting names of arguments for the visualization.
% Once bound, the \texttt{inputs} function dumps the arguments in the pickle file.
% Finally, Pytracer calls the function with the correct arguments, dumps the result, and returns it. \tristan{Summarize this paragraph in the figure caption}

%
% we can employ a case by case strategy depending on the class (i.e. numpy ufunc) or simply ignore
% the instrumentation and let the original one.
% \tristan{add a sentence to explain how callable instances are instrumented}

\begin{listing}
    \centering
\begin{lstlisting}[language=Python,style=customPython]
def get_wrapper_function(module, qualname, name, function):
    """Returns the instrumented function as a string.
    This string will be compiled with the compile builtin function

    Parameters:
        module: Name of the function module
        qualname: Qualified name of the function
        name: Name of the function
        function: Python object function

    Returns:
        wrapper_code: The code source of the wrapper
    """
    function_id = id(function)
    info = (function_id, function_module, function_qualified_name)
    wrapper_code = f"""
    def {function_wrapper_name}(*args, **kwargs):
        return generic_wrapper({info},*args,**kwargs)"""
    return wrapper_code
\end{lstlisting}
    \caption{Function to create the instrumented version of a function.
    \pytracer uses the identifier of the function instead of the 
    actual function to handle aliases and avoid duplicate instrumentation.}
    \label{fig:wrapper_creation}
\end{listing}

\begin{listing}
    \centering
\begin{lstlisting}[language=Python,style=customPython,]
def generic_wrapper(self, info, *args, **kwargs):
    """Generic wrapper saving inputs and outputs of the wrapped function.
    The id_dict dict keeps a mapping between the original function identifier
    and itself. Arguments are binded to avoid mispositioning.
    Returns the actual output of the wrapped function.

    Parameters:
        info: Tuple with the function's id, function's module and function's name
        *args: Positional calling arguments
        **kwargs: Keyword calling arguments

    Returns:
        outputs: Outputs of the wrapped function
    """
    fid, fmodule, fname = info
    function = original_function_cache.id_dict[fid]
    bind = Binding(function, *args, **kwargs)
    stack = self.backtrace()
    time = elements()
    self.inputs(time=time,
                module_name=fmodule,
                function_name=fname,
                function=function,
                args=bind.arguments,
                backtrace=stack)
    outputs = function(*bind.args, **bind.kwargs)
    self.outputs(time=time,
                 module_name=fmodule,
                 function_name=fname,
                 function=function,
                 args=outputs,
                 backtrace=stack)
    return outputs
\end{lstlisting}
    \caption{\pytracer's generic wrapper function that saves
    the inputs and outputs of the wrapped function.}
    \label{fig:generic_wrapper}
\end{listing}

\subsubsection{Instrumenting classes}

\pytracer instruments classes by wrapping their callable attributes as described previously. 
This instrumentation preserves class types, which avoids type mismatches in the instrumented application.
However, classes that describe particular base types and NumPy's universal functions (class \texttt{ufunc}) contain read-only attributes that cannot be instrumented.
By default, \pytracer returns a warning when it encounters such classes.

\subsubsection{Instrumenting class instances}

\pytracer automatically instruments class instances since it instruments classes. However, it cannot instrument or instantiate classes that have read-only attributes --- in particular class \texttt{ufunc}, --- which is problematic given their pervasiveness in scientific computing. In particular, NumPy's universal functions include
vectorized elementary mathematical and other prevalent functions. 
%\gkmod{ones}{Q: prevalent whats?}.
Universal functions are implemented in C, which prevents their modification.

To overcome this issue, \pytracer wraps the class instance into a transparent wrap class (\texttt{twc}) that overloads
1) function \texttt{\_\_getattribute\_\_}, called when an instance attribute is accessed --- instrumented to return the attributes from the original class, and 2) function \texttt{\_\_call\_\_}, used when an instance behaves like a function through the () operator --- instrumented with function \texttt{generic\_wrapper}.
% \gkmod{}{Note: the previous sentence is a bit hard to follow; the word "called" is used a lot so I got a bit lost.}
The \texttt{twc} returns the queried attribute of the instance instead of returning its attributes, allowing transparent access to the instance.  Finally, \pytracer overloads the builtin \texttt{type} function to preserve the type of the wrapped instance.

\subsubsection{Instrumenting iterators}

In functional programming, iterators traverse containers lazily, meaning that the next element in the sequence is only computed when the application uses it. This technique allows for the manipulation of virtually infinite sequences with finite memory. However, it implies that no complete mapping of the container returned by an iterator exists in memory since the application computes each element on the fly. Therefore, iterators are not serializable which implies that \pytracer cannot save them to output traces. 
A workaround would be to convert iterators into explicit containers, however, this would increase the memory footprint significantly. \pytracer, therefore, does not instrument non-builtin iterators.

% \begin{listing}
% \begin{minipage}[t]{0.4\linewidth}
%     \begin{lstlisting}[language=Python,style=customPython]
% >>> import numpy as np
% >>> x = np.array(range(10))
% >>> index = np.array([1,-1],dtype=int)
% >>> x[index]
% >>> array([1, 9])
%     \end{lstlisting}
% \end{minipage}
% \begin{minipage}[t]{0.4\linewidth}
%     \begin{lstlisting}[language=Python,style=customPython]
% >>> import numpy as np
% >>> x = np.array(range(10))
% >>> index = np.array([1,-1], dtype=object)
% >>> x[index]
% Traceback (most recent call last):
%   File "<stdin>", line 1, in <module>
% IndexError: arrays used as indices must be of integer (or boolean) type
%     \end{lstlisting}
% \end{minipage}
% \caption{Illustration of the issue of using \texttt{frompyfunc} function to convert function to \texttt{ufunc}. Left: original code. Right: instrumented code. 
% \tristan{This needs more explanation: where is the ufunc, what is the type returned by the original ufunc, and why does it crash in the instrumented version.} \Yohan{TODO}}
%     \label{fig:numpy_array_index_issue}
% \end{listing}

\subsection{Detecting numerical instabilities}

\pytracer detects numerical instabilities by computing summary statistics across multiple executions perturbed with numerical noise. While \pytracer's instrumentation, trace aggregation, and visualization work with various types of numerical noise, we experimented primarily with Monte-Carlo Arithmetic (MCA) as it is an accurate model for floating-point errors.

\subsubsection{Noise injection}
\label{sec:fuzzy}

Three  main  approaches  exist  to  detect  numerical  instabilities:  stochastic  arithmetic,  uncertainty  or  sensitivity analysis [23], and random seed analysis [11].  PyTracer is agnostic to the noise model used and therefore works for all methods mentioned.  For our use cases, we focus on stochastic arithmetic since it does not make any assumption about the noise model, in contrast to sensitivity analysis

Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors as random variables and characterize them statistically.
Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing two improvements:
(i) a virtual precision parameter, allowing to simulate reduced working precision, and (ii) different perturbation modes to introduce perturbations
on function inputs or output.
MCA replaces each floating-point number $x$ with a stochastic counterpart computed as:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
MCA can be applied in three modes: Random Rounding (RR), Precision Bounding (PB), and full MCA, which apply function $inexact$ to the output, the inputs, or both, respectively. While the RR mode is equivalent to stochastic rounding, the PB mode can also identify catastrophic cancellations.

Stochastic arithmetic quantifies numerical error using the number of significant digits $s$ estimated among the sampled values. A common formula to determine this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  
Sohier et al.~\cite{sohier2018confidence} recently provided a generalization of this formula to include confidence intervals.

We enable MCA in Python programs via Verificarlo~\cite{verificarlo}, a clang-based compiler~\cite{lattner2008llvm} that replaces floating-point operations by a generic call to a configurable floating-point model. Several floating-point models are available~\cite{chatelain2019automatic,chatelain2019outils}, also called backends.
\pytracer leverages ``fuzzy"~\cite{kiar2020comparing}, a collection of software tools compiled with Verificarlo. In particular, fuzzy provides MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK, NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub organization at \href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

% As the time of writing this article, fuzzy provides 5 levels of instrumentation (each level N including the lower levels):
% \begin{enumerate}
%     \item BLAS + LAPACK
%     \item Python interpreter
%     \item Numpy
%     \item SciPy
%     \item Scikit-learn
% \end{enumerate}

\subsubsection{Trace format}

\pytracer stores traces in the pickle format, a binary format to serialize Python objects.
The main advantages of the pickle format compared to other serialization approaches such as marshal or JSON are its ability to serialize most Python objects and its native compression.  In addition, pickle writes Python objects sequentially, which preserves the temporal ordering required by subsequent trace analyses.

Traces store numerical values at the granularity of function calls. When the application invokes a function, \pytracer saves a call input object containing contextual information (function's name, module's function, stacktrace) and the values of the function arguments. When the function returns, \pytracer saves a similar call output object containing the returned value.

\subsubsection{Trace aggregation}

Once traces are available, \pytracer sequentially parses them and computes the mean, standard deviation, and the number of significant digits (as in Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary statistics in an HDF5 file, a hierarchical format that facilitates visual browsing since only
required information are loaded when requested by the user.
% \gkmod{}{Q: Is there a reason we don't also use HDF for the raw traces? Lack of native compression?} \Yohan{Principaly because you can't save any Python objects in HDF5 while you can with Pickle}
This operation critically relies on the ordering of call input and output objects in the traces. Therefore, it assumes that the application is single-threaded and that the control flow does not depend on a random state. 
% Techniques to overcome this limitation are discussed in the discussion section.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figure/pytracer_layout.pdf}
    \caption{Pytracer visualization layout.}
    \label{fig:visu-layout}
\end{figure}

To aggregate traces, \pytracer successively unpickles call objects in each trace, checks
that the call objects refer to the same function and stacktrace, and terminates the parsing otherwise.
PyTracer stores function arguments or returned values in a NumPy array for each matching call object to compute statistics.
For compound objects such as tuples, dictionaries, or objects, it recursively parses the fields until finding
numeric types or NumPy arrays of numeric types.

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built with Plotly Dash~\cite{plotly}, a Python framework for web visualization.
Plotly Dash abstracts the low-level details of web applications in a highly customizable development interface allowing to offload heavy computations to a server if necessary.
% \gkmod{}{Q: does it support using a remote server with forwarding, in the event that your local system isn't that beefy?} \Yohan{Yes, that the meaning of the sentence. Should we reword it?}
Figure~\ref{fig:visu-layout} presents the general layout of the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
 \item Timeline view: This view displays the values of each input and output for a given function at a given invocation. By default, \pytracer shows the number of significant digits computed across the traces. The mean and standard deviation across perturbed samples are also available. 
%  \gkmod{}{Q: mean/stdev of what? the inputs/outputs? But what if they are on a totally different scale (e.g. log(0.0001)}. \Yohan{All values, a value mean one input or one output. For different scale, you can switch to the log scale. It is worth mentioning it?}
 The visualizer uses an upper triangle for inputs and a lower triangle for outputs. \pytracer internally represents all values as NumPy arrays. In the case of non-scalar values or types that cannot be trivially cast into NumPy arrays, \pytracer splits the value into several variables. Hence, \pytracer will transform a class containing $n$ floating-point parameters into $n$ variables. Each variable is prefixed with the original value and postfixed with the name of the parameter.
%  \gkmod{}{Q: how is this managed for the sake of interpretability? E.g. if one of the components has very low significance and another few are fine, how can I easily identify where in my sequence we have the drop in precision?} \Yohan{You can click on the value of interest in the legend and it will hide the others. }
    Finally, non-scalar values (vectors or matrices) are represented by their average and can be inspected with the Zoom view (see below).
 \item Gantt chart view: This view displays the call graph as a Gantt chart representing the instrumented functions. Each function call is represented with a bar.
    The time overlaps reflect the caller/callee relation.
    For example, in the figure, \texttt{function1} calls \texttt{function2} that itself calls \texttt{function4} and \texttt{function5}. 
    This view provides insights on the calling context for each function, which may influence numerical stability.
    % \gkmod{}{Note: this is a non-standard layout for gantt charts... normally the width/overlap is determined by time of start + duration. Is that information contained somewhere in here? If not, will a very fast function that calls sub-functions, say, 1000 times look longer than a very slow function that doesn't call any subfunctions?}
    % \Yohan{I don't keep the duration because the different traces may have different time execution profiles and it will not make sense to average times. Moreover, since we focus on the numerical stability and not on performance, we only care about the call order. Yes it will look longer because the unit of the X-axis is the invocation, not the time.}
 \item  Source code view: This view displays a local snapshot of the source code file of the variable hovered in the timeline view. 
 \item Zoom view: This view provide details about non-scalar values. For example, a matrix may have instabilities around its diagonal but not in other regions.
    The visualizer also provides additional metrics such as array shape, infinite norm, or condition number.
\item  Functions list view: This view allows the user to select the functions plotted in the timeline view.
\end{enumerate}

\section{Numerical evaluations}

We evaluated \pytracer on two widely-used Python libraries: SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference machine-learning framework that
 implements a wide range of state-of-the-art algorithms. For each library, we executed the software examples provided in their respective GitHub repositories. We traced each example five times, having activated MCA in the Python interpreter, BLAS, LAPACK, NumPy, SciPy, and scikit-learn. We repeated the experiment with two MCA modes, Random Rounding (RR) and Full MCA, resulting in 2 experiments for each example. We used the default virtual precision in Verificarlo, namely 24 bits for single-precision floating-point values and 53 bits for double-precision values, which simulates machine error. We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.
%  \tristan{Report scipy and sklearn versions used.}

\subsection{Results overview}

Table~\ref{tab:pytracer_results_summary} summarizes
\pytracer's outcome on SciPy and scikit-learn examples.
Columns \textit{trace} and \textit{parse} report on the tracing and aggregation steps with three possible outcomes: 1) green for success, meaning that the execution ended well and no issues were raised by the application, 2) red for numerical error in the application, meaning that the execution raised a Python error before the end, and 3) orange for errors due to the fuzzy noise injection that are not considered numerical errors, meaning that the execution raised a Python error but due to a perturbation that should not have been introduced (i.e. a floating-point value representing an integer).
% \gkmod{}{Q: What does "success" or the other labels actually mean? What is the threshold for numerical errors, e.g.}
% , like for instance index errors due to an index shift.
% \tristan{Explain what this is, in a setence here or with a subsubsection if it requires more}. 
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random Rounding mode was entirely successful for 32/40 examples and showed an average precision loss of 8 bits. The remaining examples failed due to numerical errors in the libraries, which we discuss hereafter. The instrumentation with Full MCA is more invasive and reduced the number of successful executions to 12/40. Among the 28/40 failed executions, one failed due to an actual numerical error (\texttt{Bayesian Ridge Regression} example) and the other ones failed due to 
perturbations that should not have been introduced.
% \tristan{finish this sentence}.
% \gkmod{}{Q: Does this mean the 28 failed, or completed but poorly? In general, the way the states are being referred to so far in the results are rather ambiguous}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes the precision for the main outputs of SciPy and scikit-learn examples. 
We measured the precision in significant bits on the final primary example output by taking the element-wise mean for non-scalar values and, if the example produces many outputs, the average value among all of them. The numerical quality across all examples varies from 0 to 52, with an average of 44 significant bits. For both modes, we note that only two examples failed due to a numerical error 
% \tristan{can we see that in the table? if not, don't comment on it, it is puzzling. Could you show that in the table by using the same error symbols as in Table 2?} 
(\texttt{face\_recognition} for RR and \texttt{Bayesian Ridge Regression} for Full MCA).
% even though this number is underestimated since Full MCA mode is more intrusive than RR mode.
% Indeed, most of the examples produce an error caused by the fuzzy noise injection. \gkmod{}{Note: I don't understand what is being said in the rest of the sentence following the parentheses} \Yohan{Full MCA mode perturbs values that should not be perturbed and so
% it triggers artificial bugs (as for numpy arange)}

\begin{table}[]
    \centering
    \small
    \begin{subfigure}[t]{\linewidth}
    \centering
    \begin{tabular}{|lll|c|c|c|c|c|c|}
    \hline
    \multicolumn{3}{|c}{ \multirow{2}{*}{Application} } & \multicolumn{3}{|c|}{Random Rounding (RR)} & \multicolumn{3}{c|}{Full MCA} \\
    \cline{4-9}
    & & & trace & parse & sigbits & trace & parse & sigbits \\
    \hline
    %        &               RR                 &                 MCA               \\
    %        &  numpy         &   no numpy      &      numpy      &     no numpy    \\
    %        & run   & parse  &  run   &  parse &  run   & parse  & run    & parse  \\

>>>>

\section{SciPy}
We tested \pytracer on SciPy examples provided in the tutorial website section.
SciPy is organized in several libraries targeting specific computing domains. We
focused on \texttt{fftpack} (Fast Fourier Transform routines),
\texttt{interpolate} (interpolation and smoothing splines), and
\texttt{optimize} (Optimization and root-finding routines).
